---
title: "Example workflow"
author: "Tim Daniel Rose"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

This vignette shows an example workflow for ensemble biclustering analysis 
with the `metabiclustering` package.
Every function of the package has a help page with a detailed documentation. 
To access these type `help(package=metabiclustering)` in the `R` console.

### Load packages

Import dependencies.
```{r setup, message=FALSE}
library(mosbi)
```

### Helper functions

Two additional functions are defined, to calculate z-scores of the data and 
to visualize the biclusters as a histogram.
```{r}
z_score <- function(x, margin = 2) {
    z_fun <- function(y) {
        (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE)
    }

    if (margin == 2) {
        return(apply(x, margin, z_fun))
    } else if (margin == 1) {
        return(t(apply(x, margin, z_fun)))
    }
}

bicluster_histo <- function(biclusters) {
    cols <- mosbi::colhistogram(biclusters)
    rows <- mosbi::rowhistogram(biclusters)

    graphics::par(mfrow = c(1, 2))
    hist(cols, main = "Column size ditribution")
    hist(rows, main = "Row size ditribution")
}
```
 

### 1. Download and prepare data

Biclustering will be done on a data matrix. As an example,  
lipidomics dataset from the metabolights database will be used 
`https://www.ebi.ac.uk/metabolights/MTBLS562`. 
The data consists of 40 samples (columns) and 245 lipids (rows).
```{r, message=FALSE}
# Download file
download.file(paste0("https://www.ebi.ac.uk/metabolights/ws/studies/MTBLS562/",
              "download/1fb6c4f0-e87b-463a-bc77-1a0ff345ec3d?file=",
              "m_aging_mouse_brains_metabolite_profiling_mass_",
              "spectrometry-1_v2_maf.tsv"),
    destfile = "aging_mouse_brain_data.tsv"
)

# Load data
data_table <- read.csv("aging_mouse_brain_data.tsv", sep = "\t")
data_table <- data_table[c(
    grep(
        "metabolite_identification",
        colnames(data_table)
    ),
    grep("^X", colnames(data_table))
)]

# Make data matrix
data_matrix <- z_score(log2(as.matrix(data_table[2:ncol(data_table)])), 1)

rownames(data_matrix) <- data_table$metabolite_identification

stats::heatmap(data_matrix)
```
The data has a gaussian-like distribution and no missing values, 
so we can proceed with biclustering.

### 2. Compute biclusters

The `mosbi` package is able to work with results of different 
biclustering algorithms. The  approach unites the results 
from different algorithms.
The results of four example algorithms will be computed and 
converted to `mosbi::bicluster` objects. For a list of all 
supported biclustering algorithms/packages type `?mosbi::get_biclusters` 
in the `R` console.
```{r, message=FALSE}
# Fabia
fb <- mosbi::run_fabia(data_matrix) # In case the algorithms throws an error,
# return an empty list

# isa2
BCisa <- mosbi::run_isa(data_matrix)

# Plaid
BCplaid <- mosbi::run_plaid(data_matrix)

# QUBIC
BCqubic <- mosbi::run_qubic(data_matrix)

# Merge results of all algorithms
all_bics <- c(fb, BCisa, BCplaid, BCqubic)

bicluster_histo(all_bics)
```
The histogram visualizes the distribution of bicluster sizes 
(separately for the number of rows and columns of each bicluster). 
The total number of found biclusters are given in the title.

### 3. Compute network

The next step of the ensemble approach is the computation of a 
similarity network of biclusters. To filter for for similarities due to 
random overlaps of biclusters, we apply an error 
model (For more details refer to our publication). 
Different similarity metrics are available. 
For details type `mosbi::bicluster_network` in the `R` console.
```{r}
bic_net <- mosbi::bicluster_network(all_bics, # List of biclusters
    data_matrix, # Data matrix
    n_randomizations = 5,
    # Number of randomizations for the
    # error model
    MARGIN = "both",
    # Use datapoints for metric evaluation
    metric = 4, # Fowlkes–Mallows index
    # For information about the metrics,
    # visit the "Similarity metrics
    # evaluation" vignette
    n_steps = 1000,
    # At how many steps should
    # the cut-of is evaluated
    plot_edge_dist = TRUE
    # Plot the evaluation of cut-off estimation
)
```
The two resulting plot visualize the process of cut-off estimation. 
The right plot show the remaining number of edges for the computed 
bicluster network (red) and for randomizations of biclusters (black). 
The vertical red line showed the threshold with the highest 
signal-to-noise ratio (SNR). All evaluated SNRs are again 
visualized in the left plot. 


The next plot shows the bicluster similarity matrix. 
It reveals highly similar biclusters.
```{r}
stats::heatmap(get_adjacency(bic_net))
```

#### Visualize network

Before the final step, extraction of bicluster 
communities (ensemble biclusters), the bicluster 
network can be layouted as a network.
```{r}
plot(bic_net)
```
The networks are plotted using the `igraph` package. `igraph` 
specific plotting parameters can be added.
For help type: `?igraph::plot.igraph`

To see, which bicluster was generated by which algorithm, 
the following function can be executed:
```{r}
mosbi::plot_algo_network(bic_net, all_bics, vertex.label = NA)
```


The downloaded data contains samples from different weeks of 
development. This can be visualized on the network, showing 
from which week the samples within a bicluster come from.
```{r}
# Prepare groups for plotting
weeks <- vapply(
    strsplit(colnames(data_matrix), "\\."),
    function(x) {
        return(x[1])
    }, ""
)

names(weeks) <- colnames(data_matrix)

print(sort(unique(weeks))) # 5 colors required

week_cols <- c("yellow", "orange", "red", "green", "brown")

# Plot network colored by week
mosbi::plot_piechart_bicluster_network(bic_net, all_bics, weeks,
    week_cols,
    vertex.label = NA
)
graphics::legend("topright",
    legend = sort(unique(weeks)),
    fill = week_cols, title = "Week"
)
```

Such a visualization is also possible for the samples:
```{r}
# Prepare groups for plotting
samples <- vapply(
    strsplit(colnames(data_matrix), "\\."),
    function(x) {
        return(x[2])
    }, ""
)

names(samples) <- colnames(data_matrix)

samples_cols <- RColorBrewer::brewer.pal(
    n = length(sort(unique(samples))),
    name = "Set3"
)


# Plot network colored by week
mosbi::plot_piechart_bicluster_network(bic_net, all_bics, samples,
    samples_cols,
    vertex.label = NA
)
graphics::legend("topright",
    legend = sort(unique(samples)),
    fill = samples_cols, title = "Sample"
)
```

### 4. Extract louvain communities

Calculate the communities
```{r}
coms <- mosbi::get_louvain_communities(bic_net,
    min_size = 3,
    bics = all_bics
)
# Only communities with a minimum size of 3 biclusters are saved.
```


#### Visualization of the communities
```{r}
# Plot all communities
for (i in seq(1, length(coms))) {
    tmp_bics <- mosbi::select_biclusters_from_bicluster_network(
        coms[[i]],
        all_bics
    )

    mosbi::plot_piechart_bicluster_network(coms[[i]], tmp_bics,
        weeks, week_cols,
        main = paste0("Community ", i)
    )
    graphics::legend("topright",
        legend = sort(unique(weeks)),
        fill = week_cols, title = "Week"
    )

    cat("\nCommunity ", i, " conists of results from the
             following algorithms:\n")
    cat(get_bic_net_algorithms(coms[[i]]))
    cat("\n")
}
```

#### Extraction of the communities

Finally, communities of the network can be extracted as ensemble biclusters. 
The are saved as a list of `mosbi::bicluster` objects and therefore in the 
same format as the imported results of all the algorithms. 
With the parameters `row_threshold` & `col_threshold`, 
the minimum occurrence of a row- or column-element in the biclusters 
of a community can be defined.
```{r}
ensemble_bicluster_list <- mosbi::ensemble_biclusters(coms, all_bics,
    data_matrix,
    row_threshold = .1,
    col_threshold = .1
)
```
